MLPPI Wizard: An Automated Multi-level Partitioning Tool on Analytical Workloads

### 2017-03-13 mlppi-comsis17/paper.tex
Section 2.2 updates

Whatever partition is chosen, 
the selection must be semantically correct; namely, 
the mapping of rows to partitions must be an injection. 
In other words, the constraints must form a covering of 
the entire range, so that a row will belong to exactly one and one partition. 
The Teradata optimizer then applies \hbox{partition} elimination for queries 
that specify conditions on {\tt LO\_DISCOUNT} and/or {\tt LO\_QUANTITY}. 
For \hbox{example}, only partitions~1 and~2 are needed for the query 
``{\tt SELECT * FROM LINEORDER WHERE LO\_DISCOUNT >= 7 AND} {\tt LO\_QUANTITY <= 30}.'' 
Similarly, partitions~1 and~5 are sufficient to answer the query 
``{\tt SELECT * FROM LINEORDER WHERE LO\_QUANTITY < 25}.'' 
\linebreak The respective sizes of the partitions are a factor of the data distribution. 
To see MLPPI in greater detail, check our references~\cite{klindt09mlppi,sinclair:ppi}.

@@

%Next, using {\tt LINEORDER} defined in the SSB we provide examples of partitioning options that 
%a DBA may define based on a small query set. 
\paragraph{Challenge:}
Consider a query set $Q$ having two queries $q$1 and $q$2 on the SSB, as shown below.
\vspace{-.05in}
\begin{center}
{\small
\begin{tabular}{rl}
$q$1:	& {\tt SELECT SUM(l.LO$\_$EXTENDEDPRICE*l.LO\_DISCOUNT)} \\ 
		& {\tt FROM LINEORDER l, DDATE d} \\
		& {\tt WHERE l.LO\_ORDERDATE = d.D\_DATEKEY} \\
		& {\tt AND d.D\_YEAR = `1993'} \\
		& {\tt AND l.LO\_DISCOUNT IN (1, 4, 5)} \\
        & {\tt AND l.LO\_QUANTITY <= 30} \\ 
$q$2:	& {\tt SELECT c.C\_NATION, SUM(l.LO\_REVENUE)} \\ 
		& {\tt FROM CUSTOMER c, LINEORDER l} \\
		& {\tt WHERE l.LO\_CUSTKEY = c.LO\_CUSTKEY} \\
		& {\tt AND c.C\_REGION=`EUROPE'} \\
		& {\tt AND l.LO\_DISCOUNT >= 7} \\
		& {\tt AND l.LO\_QUANTITY >= 25 AND l.LO\_QUANTITY <= 35} \\
		& {\tt GROUP BY c.C\_NATION} \\
		& {\tt ORDER BY revenue desc} \\
\end{tabular}
}
\vspace{-.01in}
\end{center}

A DBA may focus on the {\tt LINEORDER} fact table and 
the predicates involving {\tt LINEORDER} fields only. 
He can figure out the following five constraints, identified by the query number 
and the sequence number of predicates in each query: 
\vspace{-.05in}
\begin{center}
\begin{tabular}{cl}
$q$1.1: & {\tt LO\_DISCOUNT IN (1,4,5)} \\ 
$q$1.2: & {\tt LO\_QUANTITY <= 30} \\ 
$q$2.1: & {\tt LO\_DISCOUNT >= 7} \\ 
$q$2.2: & {\tt LO\_QUANTITY >= 25}\\ 
$q$2.3: & {\tt LO\_QUANTITY <= 35}. \\ 
\end{tabular}
\end{center}
\vspace{-.05in}
At this point the DBA needs to take into account options 
based only on two fields and the five constraints. 
There are many other possibilities from a {\em fine-grained} partition set 
to a {\em coarser-grained} definition.

Considering for the time being the column {\tt LO\_DISCOUNT}, 
one solution is to identify each value 
for the {\tt IN} predicate in $q$1.1 
and use $q$2.1 as is. This yields the following partitioning expression 
for {\tt LO\_DISCOUNT}:
\vspace{-.05in}
\begin{center}
{\footnotesize
\begin{tabular}{l}
{\tt CASE\_N(} \\
\hspace{0.1in}{\tt LO\_DISCOUNT = 1,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT = 4,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT = 5,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT >= 7,} \\
\hspace{0.1in}{\tt NO CASE OR UNKNOWN} \\
{)}. \\
\end{tabular}
}
\end{center}
The above expression minimizes the size of the partitions 
by focusing exactly on the values required to satisfy the constraints 
but creates a large number of small partitions.

@@

\vspace{-.1in}

@@

For each field referenced by the predicates, 
we then build a set of distinct ranges on that field. 
Specifically, we gather fields referenced by predicates in $P$, extract 
ranges from the predicates, and group the ranges of each of the referenced fields. 
Assume $I$ to be the list of fact table fields referenced by $P$. 
In the workload $Q$ we add to $I$ two fields, {\tt LO\_DISCOUNT} and {\tt LO\_QUANTITY}, used by $P$. 
The whole range set then becomes a kind of two-dimensional array, called $R$, capturing 
the range representation of the gathered predicates. 
Each entry under a field in $R$ represents a pair of (start and end) values forming the range. 
In the range representation, infinity ($\infty$) or -infinity (-$\infty$) can 
be used for unbounded ranges. 
``['' or ``]'' are used for closed ranges, and ``('' or ``)'' for open ranges. 
In particular, {\tt IN} predicate is represented by multiple ranges. 
For instance, the above $p$1 can be represented by the following three ranges 
\range{1, 1}, \range{4, 4}, \range{5, 5}; however, the last two consecutive 
ranges can be consolidated as \range{4, 5}. 
In the continuing example the field-wise distinct range set ($R$) is formed by the entries below.

@@

We then build another map, called $M$2, between $P$ and $R$. 
That is, we associate each predicate in $P$ with a corresponding range in $R$. 
Let $R$[$i$, $j$] denote the interval value for the $i$-th field and $j$-th range in $R$. 
$R$[1, 3], for instance, indicates range \ropenrange{7, \infty} under {\tt LO\_DISCOUNT}. 
In the running example, $M$2 is constructed as follows:

@@

Since there could be multiple predicates on the same field across queries, 
extracted ranges may overlap each other. 
The overlapping ranges must be broken without any common portion, so that 
a pair of consecutive, \hbox{non-overlapping} ranges can be 
considered for a merge in further phases.

Algorithm~\ref{algo:split_overlap} describes the way of breaking 
the \hbox{overlap} of ranges.

\begin{algorithm}[h]
\caption{Splitting Overlapping Ranges}
\label{algo:split_overlap}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$R$ having overlapping ranges}
\Output{$R$ with consecutive, non-overlapping ranges}
	\ForEach{field $i$ $\in$ $R$}{	
		$L$ $\leftarrow$ Get the range set of $i$ from $R$ and sort by start value \\
		$j$ = 0 \;
		\While{$j$ $<$ $|L|$-1}{
			$r_{j}$, $r_{j+1}$ $\leftarrow$ Adjacent ranges in $L$ \\
			\If{$r_{j}$.$end$ $\geq$ $r_{j+1}$.$start$}{
				Make a split by modifying values of $r_{j}$ and $r_{j+1}$. \\
				Insert into $L$ an intermediate range if any. \\
				Re-sort ranges in $L$ and reset $j$ to 0. \\
			}\lElse{$j$++;}
		}
		Update $R$ with the final $L$. 
	} 
}
\end{algorithm}

\vspace{-.1in}

\noindent For each field, the tool gets its associated ranges and sorts them 
by start value. 
It then checks whether or not adjacent ranges overlap each other.
If that is the case, then we make a split between the ranges. 
In the example, the split is applied on the 
overlapping ranges, \lopenrange{\infty, 30} and \range{25, 35}, under {\tt LO\_QUANTITY}. 
Subsequently, a common, intermediate range (\range{25, 30}) is newly created and inserted into the range set 
belonging to the {\tt LO\_QUANTITY} field, in order to fill the gap. 
Then, $R$ in the example is shown as follows.

\
{\it $R$}:
\begin{center}
\begin{tabular}{c|c} 
 {\tt LO\_DISCOUNT} & {\tt LO\_QUANTITY} \\ \hline
\range{1, 1}      & \openrange{\infty, 25} \\
\range{4, 5}      & \range{25, 30} \\
\ropenrange{7, \infty} & \range{31, 35} \\
\end{tabular}
\end{center}

The update of $R$ affects the existing map $M$2. 
In the running example, $p$3 and $p$4 influenced by the split get mapped to 
new range sets, \{$R$[2, 1], $R$[2, 2]\} and \{$R$[2, 2], $R$[2, 3]\}, respectively. 
Of course, the range sets mapped to $p$1 and $p$2 referencing {\tt LO\_DISCOUNT} remain unchanged. 
As a result, in the example we obtain the updated $M$2 as follows:

$M$2:
\begin{center}
\begin{tabular}{l}
$\langle${$p$1}, \{$R$[1, 1],$R$[1, 2]\}$\rangle$ \\
$\langle${$p$2}, \{$R$[1, 3]\}$\rangle$ \\ 
$\langle${$p$3}, \{$R$[2, 1], $R$[2, 2]\}$\rangle$ \\
$\langle${$p$4}, \{$R$[2, 2], $R$[2, 3]\}$\rangle$. \\
\end{tabular}
\end{center}

%\vspace{.1in}

@@

%The DBMS or user may impose a limit on the number of fields in the final 
%MLPPI recommendation. 
At present, the Teradata DBMS has a limit (64) of the fields 
that can be used in an MLPPI definition.
If the number of fields present in $R$ exceeds the field count limit, 
we determine which field(s) should be thrown away to satisfy the limit. 
(The star schema fact table we use has much fewer fields than the limit, 
and thus, this step will not be executed.)
To choose victim fields, the tool computes the weighted sum of 
{\em query cost} of queries regarding each field. 
The sum can be obtained by adding up every query cost on an MLPPI using 
only the ranges under the field. 
(We will cover how to measure the query cost in Section~\ref{sec:opt_phase}.) 
We incrementally discard a field with the highest query cost sum 
until the limit is reached. 
Accordingly, we can update the existing $M$2. 
One might say that the MLPPI may not be exploited if too many ranges, 
exceeding a partition count limit (or 65,536 as of now) are found in one field. 
But we assume that such an extreme case is not expected. Even if 
it happens, we can utilize the merges of consecutive ranges to make 
the MLPPI feasible.

@@

Now, the tool can create a bi-directional {\em query-to-range-set} map, called $M$, 
between $Q$ and $R$, using the existing maps $M$1 and $M$2. 
It leverages a transitive property from $M$1 to $M$2.
Once $M$ is constructed, 
the intermediate maps ($M$1 and $M$2) are not used in the rest of phases. 
$M$ is shown as below:

@@

In the subsequent phases $M$ gets used for computing costs and updated 
along with a merge of ranges, and $R$ is refined for an MLPPI recommendation.

@@

In this regard, the range set $R$ can be used to define an MLPPI using each field as one level. 
However, if the current number of partitions by $R$ is greater than 
the partition count limit, 
then it is not possible to make a feasible MLPPI based on ranges in $R$.

The actual partition count limit is ample enough to pass the running example, 
but to continue our discussion, assume that in our discussion the limit is {\it 15}.
From $R$ in the running example, we can obtain a total of (number of ranges in {\tt LO\_DISCOUNT}) 
$\cdot$(number of ranges in {\tt LO\_QUANTITY}) = (3+1) $\cdot$ (3+1) = {\it 16}
partitions. 
Because the total partitions surpass the limit, 
the tool should pass through the initial phase in which 
fewer partitions than the limit are made. 
(Otherwise, the wizard will proceed to the optimized phase immediately.) 
Note that one more range is counted per field in the calculation. 
The added range is equivalent to the ``{\tt NO CASE OR UNKNOWN)}'' case in Figure~\ref{fig:exam}.

@@

In this phase, we incrementally merge a range pair in $R$ to reduce partitions. 
The merge continues until the number of ongoing partitions 
drop below the partition count limit. 
Once reaching the limit, we can have a feasible MLPPI with 
the remaining partitions.

By and large, overall I/O cost may be increased by the merge. 
Since every row is hash-sorted in each AMP of the Teradata DBMS server, 
typically the full scan on a partition is done to retrieve 
all the rows matching a given query. 
In the case of a merged partition, we may read the non-qualifying rows 
that would not be seen before the merge, thereby paying more I/O to answer 
the query.  
To minimize the merge overhead, we pick up the range pair incurring 
the least I/O increase in a heuristic fashion.

To choose the desirable range pair for a merge, 
the tool leverages the scan cost of a query on a range pair.
Algorithm~\ref{algo:comp_scan_cost} represents how 
the scan cost can be computed on the range pair that \hbox{influences} a query. 
The scan cost represents the I/O cost to answer the query when the range pair is merged. It is 
modeled as the {\em number of blocks} that are to be read 
for the target query. To compute the scan cost of a query $q$, we use 
a corresponding scan cost query ($s$), and 
it can be constructed as follows: 
\begin{center}
$s$: {\tt SELECT * FROM $F$ WHERE} {\it CP},
\end{center} 
where $F$ is a fact table, and 
{\it CP} indicates the predicates  
restored from the ranges mapped to $q$ from $M$.

@@

If $q$ turns out to be affected by the merge of a range pair, 
then the range set associated with $q$ in $M$ is temporarily updated 
by removing the parent ranges and adding the merged one. 
The altered range set is remapped to $q$ in $M$ and then 
translated to the equivalent predicates, 
so that {\it CP} for $s$ based on the predicates can be built.
Unless the merge range pair influences $q$, then 
{\it CP} can be simply built based on the 
predicates restored from the existing ranges mapped to $q$.

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{|l|} \hline
{\tt ALTER TABLE LINEORDER} \\
{\tt MODIFY PRIMARY INDEX} \\
{\tt PARTITION BY(} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 1 AND LO\_DISCOUNT $\leq$ 5,} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 7,} \\ 
	\hspace{.2in}{\tt NO CASE OR UNKNOWN),} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_QUANTITY $<$ 25,} \\
	\hspace{.2in}{\tt LO\_QUANTITY $\geq$ 25 AND LO\_QUANTITY $\leq$ 35,} \\
	\hspace{.2in}{\tt NO CASE OR UNKNOWN)} \\ 
{\tt );} \\ \hline
\end{tabular}
\end{center}
\caption{An MLPPI Recommendation for Workload $Q$\label{fig:mlppi}}
\end{figure}

@@

@@

\section*{Acknowledgments} 
This work was carried out while the first author visited Teradata. 
The first author is now at Korea Institute of Science and Technology Information (yksuh@kisti.re.kr).
%\clearpage
%\vspace\fill

%1. Agrawal, R., Srikant, R.: Fast algorithms for mining association rules. In: Proceedings of the 20th International Conference on Very Large Databases. pp. 487–499. Morgan Kaufmann, San- tiago, Chile (1994)
%2. Bruce,K.B.,Cardelli,L.,Pierce,B.C.:Comparingobjectencodings.In:Abadi,M.,Ito,T.(eds.) Theoretical Aspects of Computer Software, Lecture Notes in Computer Science, vol. 1281, pp. 415–438. Springer-Verlag, Berlin Heidelberg New York (1997)
%3. Garcia-Molina, H., Ullman, D.J., Widom, J.: Database Systems: The Complete Book. Prentice Hall, New Jersey, USA (2002)
%4. van Leeuwen, J. (ed.): Computer Science Today. Recent Trends and Developments, Lecture Notes in Computer Science, vol. 1000. Springer-Verlag, Berlin Heidelberg New York (1995)
%5. Ribicˇre,M.,Charlton,P.:Ontologyoverview.MotorolaLabs,Paris(2002),[Online].Available:
%http://www.fipa.org/docs/input/f-in-00045/f-in-00045.pdf (current October 2003)
%6. Wang,X.,Bettini,C.,Brodsky,A.,Jajoida,S.:Logicaldesignfortemporaldatabaseswithmul-
%tiple granularities. ACM Transactions on Database Systems 22(2), 115–170 (1997)

%\begin{thebibliography}{99}
%\expandafter\ifx\csname url\endcsname\relax
%  \def\url#1{\texttt{#1}}\fi
%\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
%\expandafter\ifx\csname href\endcsname\relax
%  \def\href#1#2{#2} \def\path#1{#1}\fi
%
%{
%\bibitem{Suh12}
%{Y.-K.~Suh, A.~Ghazal, A.~Crolotte, and P.~Kostamaa}, {A New Tool for Multi-level Partitioning in Teradata}, 
%in: CIKM'12, pp. 2214--2218.
%  
%\bibitem{sinclair:ppi}
%P.~Sinclair, {Using PPIs to Improve Performance},
%  \url{http://www.teradata.com/tdmo/v08n03/pdf/AR5731.pdf}
%  (accessed August 27, 2015).
%
%\bibitem{klindt09mlppi}
%J.~Klindt, {Single-level and Multilevel Partitioned Primary \linebreak Indexes},
%  \url{http://www.teradata.com/white-papers/Single-level-and-}\linebreak\url{Multilevel-Partitioned-Primary-Indexes-eb1889/}
%  (accessed July 27 2015).
%
%\bibitem{oneil:ssb}
%{P.~O’Neil, B.~O'Neil, and X.~Chen}, {The Star Schema Benchmark (SSB)},
%  \url{http://www.cs.umb.edu/~poneil/StarSchemaB.PDF} 
%  (accessed February 28, 2017).
%
%\bibitem{agrawal04:integrating}
%{S.~Agrawal, V.~Narasayya, and B.~Yang}, {Integrating Vertical and Horizontal Partitioning into Automated Physical Database Design}, 
%in: \hbox{SIGMOD} '04, pp. 359--370.
%
%\bibitem{Lightstone04:db2auto}
%{S.~Lightstone and B.~Bhattacharjee}, {Automated Design of
%  Multi-dimensional Clustering Tables for Relational Databases}, in:
%  VLDB '04, pp. 1170--1181.
%
%\bibitem{nehme2011automated}
%{R.~Nehme and N.~Bruno}, {Automated Partitioning Design in Parallel
%  Database Systems}, in: SIGMOD '11, pp. 1137--1148.
%
%\bibitem{chaudhuri1998autoadmin}
%{S.~Chaudhuri and V.~Narasayya}, {AutoAdmin `What-If' Index Analysis Utility}, ACM SIGMOD Rec. 27~(2) (1998) 367--378.
%
%\bibitem{finkelstein1988physical}
%{S.~Finkelstein, M. Schkolnick, and P. Tiberio}, {Physical Database Design for Relational Databases}, ACM Trans. on Datab. Syst. (TODS), 13~(1) (1988) 91--128.
%
%\bibitem{labio1997physical}
%{W.~Labio, D.~Quass, and B.~Adelberg}, {Physical Database Design for Data Warehouses}, in: ICDE '97, pp. 277--288.
%
%\bibitem{rozen1991framework}
%{S.~Rozen and D.~Shasha}, {A Framework for Automating Physical Database Design}, in: VLDB '91, pp. 401--411.
%
%\bibitem{Zilio1998}
%{C.~Zilio}, {Physical Database Design Decision Algorithms and Concurrent
%  \hbox{Reorganization} for Parallel Database Systems}, Ph.D. thesis,
%  Dept. of CS, Univ. of Toronto (1998).
%
%\bibitem{sheet2009oracle}
%{Oracle}, {Oracle White Paper---Partitioning with \hbox{Oracle} Database 11g
%  \hbox{Release 2}},
%  \url{http://www.oracle.com/technetwork/database/options/partitioning/twp-partitioning-11gr2-2009-09-130569.pdf}
%  (accessed June 23, 2014).
%
%\bibitem{saa2007oracle}
%{Oracle}, {SQL Access Advisor},
%  \url{http://docs.oracle.com/cd/B19306_01/server.102/b14211/advisor.htm}
%  (accessed March 30, 2012).
%
%\bibitem{rao2002automating}
%{Jun Rao, Chun Zhang, Nimrod Megiddo, and Guy Lohman}, {Automating
%  Physical Database Design in A Parallel Database}, in: SIGMOD '02, pp. 558--569.
%
%\bibitem{agrawal2000automated}
%{S.~Agrawal et. al.}, {Automated Selection of Materialized Views and Indexes in SQL Databases}, in:
%  {VLDB '00}, pp. 496--505.
%
%\bibitem{Agrawal04:sqlserver}
%{S.~Agrawal et. al.}, {Database Tuning Advisor for Microsoft
%  SQL Server 2005: Demo}, in: \hbox{SIGMOD} '05, pp. 930--932.
%
%\bibitem{dash2011cophy}
%{D.~Dash}, {CoPhy: A
%  Scalable, Portable, and Interactive Index Advisor for Large Workloads},
%  in: PVLDB, 4~(6) (2011) 362--372.
%
%\bibitem{Kimura11}
%{H.~Kimura et. al.}, {Compression Aware

>>>>

Whatever partition is chosen, 
the selection must be semantically correct; namely, 
the mapping of rows to partitions must be an injection. 
In other words, the constraints must form a covering of 
the entire range, so that a row will belong to exactly one and one partition. 
The Teradata optimizer then applies \hbox{partition} elimination for queries 
that specify conditions on {\tt LO\_DISCOUNT} and/or {\tt LO\_QUANTITY}. 
For \hbox{example}, only partitions~1 and~2 are needed for the query 
``{\tt SELECT * FROM LINEORDER WHERE} \linebreak {\tt LO\_DISCOUNT >= 7 AND} {\tt LO\_QUANTITY <= 30}.'' 
Similarly, partitions~1 and~5 are sufficient to answer the query 
``{\tt SELECT * FROM LINEORDER WHERE} \linebreak {\tt LO\_QUANTITY < 25}.'' 
The respective sizes of the partitions are a factor of the data distribution. 
To see MLPPI in greater detail, check our references~\cite{klindt09mlppi,sinclair:ppi}.

@@

%Next, using {\tt LINEORDER} defined in the SSB we provide examples of partitioning options that 
%a DBA may define based on a small query set. 
\paragraph{Challenge:}
Consider the below query set $Q$ having two queries $q$1 and $q$2 on the SSB.
\vspace{-.05in}
\begin{center}
{\small
\begin{tabular}{rl}
$q$1:	& {\tt SELECT SUM(l.LO$\_$EXTENDEDPRICE*l.LO\_DISCOUNT)} \\ 
		& {\tt FROM LINEORDER l, DDATE d} \\
		& {\tt WHERE l.LO\_ORDERDATE = d.D\_DATEKEY} \\
		& {\tt AND d.D\_YEAR = `1993'} \\
		& {\tt AND l.LO\_DISCOUNT IN (1, 4, 5)} \\
        & {\tt AND l.LO\_QUANTITY <= 30} \\ 
$q$2:	& {\tt SELECT c.C\_NATION, SUM(l.LO\_REVENUE)} \\ 
		& {\tt FROM CUSTOMER c, LINEORDER l} \\
		& {\tt WHERE l.LO\_CUSTKEY = c.LO\_CUSTKEY} \\
		& {\tt AND c.C\_REGION=`EUROPE'} \\
		& {\tt AND l.LO\_DISCOUNT >= 7} \\
		& {\tt AND l.LO\_QUANTITY >= 25 AND l.LO\_QUANTITY <= 35} \\
		& {\tt GROUP BY c.C\_NATION} \\
		& {\tt ORDER BY revenue desc} \\
\end{tabular}
}
\vspace{-.01in}
\end{center}

A DBA may focus on the {\tt LINEORDER} fact table and 
the predicates involving \linebreak {\tt LINEORDER} fields only. 
He can figure out the following five constraints, identified by the query number 
and the sequence number of predicates in each query: 
\vspace{-.05in}
\begin{center}
\begin{tabular}{cl}
$q$1.1: & {\tt LO\_DISCOUNT IN (1,4,5)} \\ 
$q$1.2: & {\tt LO\_QUANTITY <= 30} \\ 
$q$2.1: & {\tt LO\_DISCOUNT >= 7} \\ 
$q$2.2: & {\tt LO\_QUANTITY >= 25}\\ 
$q$2.3: & {\tt LO\_QUANTITY <= 35}. \\ 
\end{tabular}
\end{center}
\vspace{-.05in}
At this point the DBA needs to take into account options 
based only on two fields and the five constraints. 
There are many other possibilities from a {\em fine-grained} partition set 
to a {\em coarser-grained} definition.

Considering for the time being the column {\tt LO\_DISCOUNT}, 
one solution is to identify each value 
for the {\tt IN} predicate in $q$1.1 
and use $q$2.1 as is. This yields the following partitioning expression 
for {\tt LO\_DISCOUNT}:
\vspace{-.05in}
\begin{center}
{\footnotesize
\begin{tabular}{l}
{\tt CASE\_N(} \\
\hspace{0.1in}{\tt LO\_DISCOUNT = 1,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT = 4,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT = 5,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT >= 7,} \\
\hspace{0.1in}{\tt NO CASE OR UNKNOWN} \\
{)}. \\
\end{tabular}
}
\end{center}
This expression minimizes the size of the partitions 
by focusing exactly on the values required to satisfy the constraints 
but creates a large number of small partitions.

@@

@@

Now we build a set of distinct ranges for each field referenced by the predicates. 
Specifically, we gather fields referenced by predicates in $P$, extract 
ranges from the predicates, and group the ranges of each of the referenced fields. 
Suppose $I$ to be a list of fact table fields referenced by $P$. 
In the workload $Q$ we add to $I$ two fields, {\tt LO\_DISCOUNT} and {\tt LO\_QUANTITY}, used by $P$. 
The whole range of the gathered predicates can be represented by a kind of two-dimensional array, called $R$.

Each entry under a field in $R$ represents a pair of (start and end) values forming the range. 
When it comes to range representation for an entry, 
``['' and ``]'' are used for closed ranges while ``('' and ``)'' for open ranges. 
Also, infinity ($\infty$) and -infinity (-$\infty$) can be used for unbounded ranges. 
In particular, {\tt IN} predicate is represented by multiple ranges. 
For instance, the above $p$1 can be represented by the following three ranges 
\range{1, 1}, \range{4, 4}, \range{5, 5}; however, the last two consecutive 
ranges can be consolidated as \range{4, 5}. 
In the continuing example the field-wise distinct range set ($R$) is formed by the entries below.

@@

In turn, we build another map, called $M$2, between $P$ and $R$. 
That is, we associate each predicate in $P$ with a corresponding range in $R$. 
Let $R$[$i$, $j$] denote the interval value for the $i$-th field and $j$-th range in $R$. 
$R$[1, 3], for instance, indicates range \ropenrange{7, \infty} under {\tt LO\_DISCOUNT}. 
In the running example, $M$2 is constructed as follows:

@@

Since there may be multiple predicates on the same field across queries, 
extracted ranges for the field may overlap each other. 
The overlapping ranges must be broken without any common portion, in order 
that we can consider a pair of consecutive, \hbox{non-overlapping} ranges for 
a merge in subsequent phases.

\begin{algorithm}[t]
\caption{Splitting Overlapping Ranges}
\label{algo:split_overlap}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$R$ having overlapping ranges}
\Output{$R$ with consecutive, non-overlapping ranges}
	\ForEach{field $i$ $\in$ $R$}{	
		$L$ $\leftarrow$ Get the range set of $i$ from $R$ and sort by start value \\
		$j$ = 0 \;
		\While{$j$ $<$ $|L|$-1}{
			$r_{j}$, $r_{j+1}$ $\leftarrow$ Adjacent ranges in $L$ \\
			\If{$r_{j}$.$end$ $\geq$ $r_{j+1}$.$start$}{
				Make a split by modifying values of $r_{j}$ and $r_{j+1}$. \\
				Insert into $L$ an intermediate range if any. \\
				Re-sort ranges in $L$ and reset $j$ to 0. \\
			}\lElse{$j$++;}
		}
		Update $R$ with the final $L$. 
	} 
}
\end{algorithm}

Algorithm~\ref{algo:split_overlap} describes the way of breaking 
the \hbox{overlap} of ranges. 
For each field, the algorithm 
gets associated ranges of each field and sorts them by start value. 
We then check whether or not adjacent ranges overlap each other.
If that is the case, we make a split between the ranges. 
In the example, the split is applied on the 
overlapping ranges, \lopenrange{\infty, 30} and \range{25, 35} in {\tt LO\_QUANTITY}. 
Subsequently, a common (or intermediate) range (\range{25, 30}) is created and inserted into the range set 
belonging to the {\tt LO\_QUANTITY} field, in order to fill the gap. 
$R$ is shown in the following.

{\it $R$}:
\begin{center}
\begin{tabular}{c|c} 
 {\tt LO\_DISCOUNT} & {\tt LO\_QUANTITY} \\ \hline
\range{1, 1}      & \openrange{\infty, 25} \\
\range{4, 5}      & \range{25, 30} \\
\ropenrange{7, \infty} & \range{31, 35} \\
\end{tabular}
\end{center}

The update of $R$ affects the existing map $M$2. 
In the running example, $p$3 and $p$4 influenced by the split get mapped to 
new range sets, \{$R$[2, 1], $R$[2, 2]\} and \{$R$[2, 2], $R$[2, 3]\}, respectively. 
Of course, the range sets corresponding to $p$1 and $p$2 referencing {\tt LO\_DISCOUNT} remain unchanged. 
As a result, in the running example we obtain the updated $M$2 in the following.

\pagebreak
$M$2:
\begin{center}
\begin{tabular}{l}
$\langle${$p$1}, \{$R$[1, 1],$R$[1, 2]\}$\rangle$ \\
$\langle${$p$2}, \{$R$[1, 3]\}$\rangle$ \\ 
$\langle${$p$3}, \{$R$[2, 1], $R$[2, 2]\}$\rangle$ \\
$\langle${$p$4}, \{$R$[2, 2], $R$[2, 3]\}$\rangle$. \\
\end{tabular}
\end{center}

@@

One question that occurs is what if a DBMS has 
a limit of fields that can be used for a partitioning definition. 
(In the case of Teradata DBMS, up to 64 fields can be allowed for 
an MLPPI definition.)
If the number of fields present in $R$ exceeds the field count limit, 
we determine which field(s) should be thrown away to satisfy the limit. 
(The star schema fact table we use has much fewer fields than the limit, 
and thus, this step will not be executed.)
To choose victim fields, our tool computes the weighted sum of 
{\em query cost}, to be discussed in Section~\ref{sec:opt_phase}, 
of queries regarding each field. 
The sum can be obtained by adding up every query cost on an MLPPI using 
only the ranges under the field. 
We incrementally discard a field with the largest sum of query cost 
until the limit is reached. 
Correspondingly, we can update the existing $M$2. 
One might say that such an MLPPI may not be exploited if too many ranges 
over a partition count limit (65,536 in Teradata DBMS) are found in one field.
But we assume that such an extreme case is not expected. Even if 
such a case happens, the merges of consecutive ranges can make the MLPPI feasible.

@@

Using the existing maps $M$1 and $M$2, 
the tool can create the final bi-directional {\em query-to-range-set} map, say $M$, 
between $Q$ and $R$. 
To construct $M$, we leverage a transitive property from $M$1 to $M$2.
Once we compute $M$, there is no need to keep 
the intermediate maps ($M$1 and $M$2) for the subsequent phases. 
In the running example, we can build $M$ in the following.

\pagebreak

@@

\noindent In the subsequent phases $M$ gets used for computing costs and updated 
along with a merge of ranges, and $R$ is refined for an MLPPI recommendation.

@@

The range set $R$ can be used to define an MLPPI using each field as one level. 
However, if the current number of partitions by $R$ is greater than 
the aforementioned partition count limit, 
then it is not possible to make a feasible MLPPI based on the ranges in $R$.

The actual partition count limit is ample enough to pass the running example, 
but to continue our discussion, assume that in our discussion the limit is {\it 15}.
From $R$ in the running example, we obtain a total of (number of ranges in {\tt LO\_DISCOUNT}) 
$\cdot$(number of ranges in {\tt LO\_QUANTITY}) = (3+1) $\cdot$ (3+1) = {\it 16}
partitions. Note here that one reserved range covering the rest but the identified 
is added by default to each field in the calculation, 
and the default range is equivalent to the ``{\tt NO CASE OR UNKNOWN)}'' case in Figure~\ref{fig:exam}. 
Since the total partitions surpass the limit, 
the tool should go through the initial phase in which 
we make fewer partitions than the limit. 
(Otherwise, the wizard can bypass the initial phase and then immediately proceed to the optimized phase.)

@@

This phase incrementally merges a range pair in $R$ to reduce partitions. 
The merge continues until the number of ongoing partitions 
drops below the partition count limit. 
After finishing the merge, the tool can derive a feasible MLPPI with the remaining partitions.

By and large, overall I/O cost may be increased by the merge, 
in that the full scan on a partition is performed to retrieve all the rows potentially matching a given query. 
In the case of a merged partition, we may read the non-qualifying rows 
that would not be seen before the merge, thereby paying more I/O to answer 
the query. To minimize the merge overhead, we pick up the range pair incurring 
the least I/O increase in a heuristic fashion. 
To choose the desirable range pair for a merge, 
the tool leverages the {\em scan cost} of a query on a range pair.

Algorithm~\ref{algo:comp_scan_cost} represents how 
the scan cost can be computed on the range pair that \hbox{influences} a given query. 
The scan cost represents the I/O cost to answer the query when the range pair is merged. 
It is defined as the {\em number of blocks} that are to be read 
for the given query. To compute the scan cost of a query $q$, we write 
a corresponding scan cost query ($s$), which can be constructed as follows: 
\begin{center}
$s$: {\tt SELECT * FROM $F$ WHERE} {\it CP},
\end{center} 
where $F$ is a fact table, and 
{\it CP} indicates a predicate(s) reconstructed from 
a range(s) mapped to $q$ in $M$.

@@

If $q$ turns out to be affected by the merge of a range pair, 
then the range set associated with $q$ in $M$ is temporarily updated 
by removing the parent ranges and adding the merged range. 
The altered range set is remapped to $q$ in $M$ and then 
translated to the equivalent predicates. Then {\it CP} for $s$ can be built based on the predicates.
Unless the merge range pair influences $q$, 
we can simply derive {\it CP} from the existing ranges mapped to $q$.

@@

@@

\begin{figure}[h]
\begin{center}
\begin{tabular}{|l|} \hline
{\tt ALTER TABLE LINEORDER} \\
{\tt MODIFY PRIMARY INDEX} \\
{\tt PARTITION BY(} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 1 AND LO\_DISCOUNT $\leq$ 5,} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 7,} \\ 
	\hspace{.2in}{\tt NO CASE OR UNKNOWN),} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_QUANTITY $<$ 25,} \\
	\hspace{.2in}{\tt LO\_QUANTITY $\geq$ 25 AND LO\_QUANTITY $\leq$ 35,} \\
	\hspace{.2in}{\tt NO CASE OR UNKNOWN)} \\ 
{\tt );} \\ \hline
\end{tabular}
\end{center}
\caption{An MLPPI Recommendation for Workload $Q$\label{fig:mlppi}}
\end{figure}

@@

\section*{Acknowledgments} 
This work was carried out while the first author visited Teradata. 
The first author is now at Korea Institute of Science and Technology Information (yksuh@kisti.re.kr).
